A Graph Theoretic Approach to Problem Formulation for Multidisciplinary Design Analysis and Optimization

### 2013-09-11 graph_syntax.tex
addresses second reviewers comments
Update with BG and DP edits to paper and reviewer response

\section{Graph-Based Syntax Definition and Formalisms}
  \label{s:syntax definition}
  In this section we present our graph-based syntax, beginning with the necessary graph 
  theoretic fundamentals to construct the graphs discussed in Sec.~\ref{s:requirements}. 
  The notation used in this paper is adapted from Diestel \cite{Diestel2010}. 
  A \emph{graph} is a pair $G = (V,E)$ of sets such that $E \subseteq V \times V$, 
  which means that the elements of $E$ are 2-element subsets of $V$. The set $V$ 
  contains the \emph{vertices} or \emph{nodes} and the set $E$ contains the \emph{edges}.
  For a \emph{directed graph} (or \emph{digraph}) we construct $E$ as a set of ordered pairs instead 
  of a set of sets. Each ordered pair represents an edge starting at the node 
  indicated by the first entry and directed to the node indicated by the second 
  entry. An edge $e$ = $(x,y)$ may simply be referred to as $xy$. The edges 
  directed out from node $v$ are denoted by $E(v)$ and the edges directed into $v$ are denoted 
  by $E^{-1}(v)$. $E(v)$ may be the empty set, a single edge, or a set of edges, and likewise for $E^{-1}(v)$.
  \begin{figure}[htb!]
    \begin{center}
    \includegraphics[width=1.5in]{images/example_directed_graph}
    \end{center}
    \vspace{-20pt}
  \caption{Example directed graph.}
  \label{f:example directed graph}
  \end{figure}
  As an example, for the directed graph shown in Fig.~\ref{f:example directed graph} we have
  \begin{IEEEeqnarray*}{rCl}
  V & = & \{1,2,3,4,5\}, \\
  E & = & \big\{(2,1),(3,2),(1,3),(3,4)\big\}.
  \end{IEEEeqnarray*}

@@

  Lastly, the cardinality of a set $B$ is the number of elements in $B$ and is denoted as $|B|$.

@@

  \begin{description}
  \item[\bf{connection edge:}] represents exchange of information external to analysis tools,
  \item[\bf{function edge:}] represents exchange of information internal to analysis tools,
  \item[\bf{driven edge:}] represents passing of information from a driver node to a 
  variable node (present only in a PSG).
  \end{description}
  Figure \ref{f:sellar types} demonstrates the usage of these node and edge 
  types via the Sellar problem. 
In this figure, function nodes are indicated by squares, 
  variable nodes are indicated by circles, connection edges are indicated by 
  dotted lines, and function edges are indicated by solid lines.
  \begin{figure}[htb!]
    \begin{center}
      \includegraphics[width=3.25in]{images/sellar_types}
    \end{center}
    \caption{Sellar problem represented as a graph.}
  \label{f:sellar types}
  \end{figure}

@@

\subsection{Analysis Blocks}
\label{ss:analysis blocks}
  Analysis tools take in a set of input variables and calculate 
  the values for their respective outputs. We represent this process
  by a digraph called an \emph{analysis block}; a notional analysis block is shown 
  in Fig.~\ref{f:analysis block}. 
  %Each analysis block is a directed graph denoted as $ A=(V_{A},E_{A})$
  As indicated in this figure, analysis blocks comprise three sets of nodes 
  representing the distinct local inputs, the analysis (computation), and the 
  distinct local outputs. The local input and local output nodes are variable 
  nodes, while the nodes representing the analysis are function nodes. All of 
  the edges within the analysis block are function edges, and are considered 
  fixed to the analysis block. This graph structure demonstrated by 
  Fig.~\ref{f:analysis block} satisfies the rules listed in Sec.~\ref{ss:types}. 
  Conversely, the rules ensure that analysis tools are represented via this analysis 
  block structure.

@@

  \subsection{Driver Nodes and Driven Edges}
  Within this syntax, all iterative processes are represented with Driver nodes. 
  The driver node shares some basic qualities with the model node. It can have incoming 
  and outgoing function edges to and from variable nodes. For example, many optimizers take 
  convergence tolerance as input and output iteration counts and final objective value. 
  With respect to variable nodes and function edges, drivers behave exactly the same as
  model nodes and are subject to the same rules governing their use in the graph. In this
  context, driver nodes can be part of driver blocks which behave just like analysis 
  blocks.

  Driven edges are the distinguishing characteristic associated with driver nodes and
  driver blocks. Driven edges don't follow the same rules as other edges. Driven edges  
  will always connect a driver node to a variable node (in either direction), but they 
  are not considered when counting the indegree or outdegree of the variable node.
  Hence a single variable node could have multiple incoming driven edges from different 
  driver blocks. Figure \ref{f:driver block} shows a notional example where a driver block 
  represents and optimizer solving an unconstrained minimization problem.

  \begin{figure}[htb]
    \begin{center}
    \includegraphics[width=3.0in]{images/driver_block}
    \end{center}
    \vspace{-10pt}
  \caption{Example driver block with node types labeled.}
  \label{f:driver block}
  \end{figure}

@@

\subsection{Design Variables, Holes, and Collisions}
  Design variables are input variables whose values are free to be 
  changed by the designer or by an optimizer in an MDAO study. In the graph-based 
  syntax, a node corresponds to a design variable if the indegree limits of the node are 
  prescribed as zero.

  If the indegree limits are violated by the number of incoming connection edges 
  (the indegree of the node), then the graph is regarded as an invalid problem formulation. 
  There are two classifications for these violations:
    \begin{description}
      \item[\bf{hole:}] The number of incoming edges is less than the lower indegree limit:
        \begin{equation} 
          \txt{deg}^-(v) < \txt{deg}_l^-(v). 
          \label{e:hole} 
        \end{equation}
        A hole represents a lack of information being supplied to the variable node. 
        This implies that the analysis tool being represented by the analysis block 
        would not be capable of producing outputs.
      \item[\bf{collision:}] The number of incoming edges is greater than the upper 
      indegree limit:
        \begin{equation} 
        \txt{deg}^-(v) > \txt{deg}_u^-(v). 
        \label{e:collision}
        \end{equation}A collision represents redundant information being supplied to 
        the variable node from two or more sources. This conflict implies an 
        ambiguity as to which information is to be used as an input for the analysis tool.
	In multifidelity problems (discussed subsequently), the upper indegree limit could be 
  specified as $deg_u^- > 2$, which implies that a collision is not noted in the case of 
  multiple inputs.
  \end{description}

@@

\subsection{Coupling Between Analyses}
  In MDAO, coupling is the mutual dependence between two or more analysis 
  tools and their respective outputs. In this graph-based syntax, coupling is 
  described by the presence of a cycle between two or more analysis blocks. In the Sellar 
  problem from Eq.~\ref{eqn:simple_fpf} 
  a mutual dependence between $D_1$ and $D_2$ exists through the variables $y_1$ and $y_2$. 
  The edges belonging to a cycle in the Sellar problem are highlighted in 
  Fig. \ref{f:sellar cycles}.  These edges collectively form a closed path between 
  analysis blocks $D_1$ and $D_2$.

@@

  Coupling cycles do not contain driver nodes in the MCG or FPG representation. 
  Solvers, optimizers, and other iterative processes are not invovled in the coupling 
  definition; however, one or more solvers will be required to build a valid PSG from a 
  given FPG  that includes cycles. Additionally, a coupling cycle has no inherent 
  start or end point. It would be acceptable to select a
  node in the cycle as a starting point and proceed around the
  loop until the starting point is reached again. For the Sellar Problem, selecting 
  $D_1$ as the starting point would yield a problem as given in 
  Eq.~\ref{eqn:simple}.
  Larger problems can involve more complex cycles in their FPG, indicating more 
  complex coupling between analyses. For example, a cycle can involve more than 
   two analysis blocks. Multiple independent cycles could also exist, indicating 
  independent coupling relationships. Cycles can also overlap, meaning that the same analysis 
  blocks are involved in multiple different coupling cycles. All of these situations
  arise naturally as the size of problems grows, and managing this coupling may
  become difficult. In the present work, Sec.~\ref{s:example problem}
  demonstrates how building an FPG from an MCG provides an opportunity to 
  identify and potentially reduce the number of cycles in a graph.

@@

\subsection{Multi-fidelity Problems}
  \label{ss:multi-fideliy problems}
  A multi-fidelity MDAO problem can be characterized by a formulation in which 
  two or more different analyses each calculate the same data. Multi-fidelity 
  is represented in a graph by a variable node having an indegree greater than 
  one, which means that multiple connection edges are directed into it. When the 
  upper indegree limit of a variable node is set above one, then the node is 
  implied to allow multiple fidelity instances of the variable. When the lower 
  indegree limit of a variable node is set above one, then multiple fidelity instances are required.

@@

  Multi-fidelity problems are always characterized by graphs in which at least some of the variable 
  nodes that have an $\txt{deg}^-_u(v) > 1$. These problems require 
  special techniques for resolving the conflicting edges by introducing some mechanism
  to manage when each of the different fidelity analyses are 
  run \cite{march2012provably,alexandrov2001approximation,Huang_Allen_Notz_Miller_2006}.
  The specifics of this mechanism are not represented within an MCG or 
  FPG. Instead, the multi-fidelity mechanism specifics would be represented as a 
  driver in a PSG derived from a multi-fidelity FPG.

Table \ref{t:variable node classification} summarizes the classification of a variable 
node as a hole, design variable, single valid input (nominal), collision, or 
multi-fidelity node.

\begin{table}[htbp]
  \centering
  \caption{Variable Node Classification}
    \begin{tabular}{ccccc}
    \toprule
    $\txt{deg}^-_l(v)$ & $\txt{deg}^-_u(v)$ & $\txt{deg}^-(v)$ & valid & classification \\
    \midrule
    0     & 0     & 0     & yes   & design variable \\
    0     & 0     & 1     & yes   & collision \\
    0     & 1     & 0     & yes   & design variable \\
    0     & 1     & 1     & yes   & single input \\
    1     & 1     & 0     & no    & hole \\
    1     & 1     & 1     & yes   & single input \\
    1     & 1     & $>1$ & no    & collision \\
    1     & 2     & 1     & yes   & supplied input \\
    1     & 2     & 2     & yes   & multi-fidelity \\
    2     & 2     & $<2$ & no    & hole \\
    2     & 2     & 2     & yes   & multi-fidelity \\
    2     & 2     & $>2$     & no   & collision \\
    \bottomrule
    \end{tabular}%
  \label{t:variable node classification}%
\end{table}%

\subsection{Iterative Control: Drivers}
  There are many iterative processes that are used to help solve MDAO problems such as 
  solvers, optimizers, and Design of Experiment (DOE). Abstractly, even a human picking 
  values guided by experience and trend could be viewed as an iterative process. 
  These processes represent the control structures that govern how a design problem 
  marches toward a solution but they do not fundamentally impact the definition of the 
  problem to be solved. e.g. which optimizer you choose does not change the problem you 
  are solving but will change the path you follow to solve it. Hence, as a rule, driver 
  nodes and driven edges are excluded from both the MCG and FPG which deal exclusively 
  with problem formulation. They only come into play with in the PSG where the issue of 
  how to solve a specific problem is addressed. In fact, at least one driver block is 
  required to be present in any PSG in order to select values for the design variables.

>>>>

\section{Graph-Based Syntax Definition and Formalisms}
  \label{s:syntax definition}
  In this section we present our graph-based syntax, beginning with the necessary graph 
  theoretic fundamentals to construct the graphs discussed in Sec.~\ref{s:requirements}. 
  The notation used in this paper is adapted from Diestel \cite{Diestel2010}. 
  A \emph{graph} is a pair $G = (V,E)$ of sets such that $E \subseteq V \times V$, 
  which means that the elements of $E$ are 2-element subsets of $V$. The set $V$ 
  contains the \emph{vertices} or \emph{nodes} and the set $E$ contains the \emph{edges}.
  For a \emph{directed graph} (or \emph{digraph}) we construct $E$ as a set of ordered pairs instead 
  of a set of sets. Each ordered pair represents an edge starting at the node 
  indicated by the first entry and directed to the node indicated by the second 
  entry. The edges 
  directed out from node $v$ are denoted by $E(v)$ and the edges directed into $v$ are denoted 
  by $E^{-1}(v)$. $E(v)$ may be the empty set, a single edge, or a set of edges, and likewise for $E^{-1}(v)$.
  \begin{figure}[htb!]
    \begin{center}
    \includegraphics[width=1.5in]{images/example_directed_graph}
    \end{center}
    \vspace{-20pt}
  \caption{Example directed graph.}
  \label{f:example directed graph}
  \end{figure}
  As an example, for the directed graph shown in Fig.~\ref{f:example directed graph}, we have
  \begin{IEEEeqnarray*}{rCl}
  V & = & \{1,2,3,4,5\}, \\
  E & = & \big\{(2,1),(3,2),(1,3),(3,4)\big\}.
  \end{IEEEeqnarray*}

@@

  Lastly, the cardinality of a set $B$ is the number of elements in $B$ and is denoted as $|B|$, and the set difference is denoted with a backslash, i.e. $A \setminus B = \{x \in A \st x \notin B\}$.

@@

  \begin{description}
  \item[\bf{connection edge:}] represents exchange of information external to analysis tools,
  \item[\bf{function edge:}] represents exchange of information internal to analysis tools,
  \item[\bf{driven edge:}] represents passing of information between a driver node and a 
  variable node (present only in a PSG).
  \end{description}
  Figure \ref{f:sellar types} demonstrates the usage of these node and edge 
  types via the Sellar problem. 
In this figure, function nodes are indicated by squares, 
  variable nodes are indicated by circles, connection edges are indicated by 
  dotted lines, and function edges are indicated by solid lines.
  \begin{figure}[htb!]
    \begin{center}
      \includegraphics[width=3.25in]{images/sellar_types}
    \end{center}
    \caption{Sellar problem represented as a graph.}
  \label{f:sellar types}
  \end{figure}

@@

\subsection{Analysis Blocks}
\label{ss:analysis blocks}
  Analysis tools accept a set of input variables and calculate 
  the values for their respective outputs. We represent this process
  by a digraph called an \emph{analysis block}; a notional analysis block is shown 
  in Fig.~\ref{f:analysis block}. 
  %Each analysis block is a directed graph denoted as $ A=(V_{A},E_{A})$
  As indicated in this figure, analysis blocks comprise three sets of nodes 
  representing the distinct local inputs, the analysis (computation), and the 
  distinct local outputs. The local input and local output nodes are variable 
  nodes, while the nodes representing the analysis are function nodes. All of 
  the edges within the analysis block are function edges, and are considered 
  fixed to the analysis block. The graph structure demonstrated by 
  Fig.~\ref{f:analysis block} satisfies the rules listed in Sec.~\ref{ss:types}. 
  Conversely, the rules ensure that analysis tools are represented via this analysis 
  block structure.

@@

  \subsection{Driver Nodes and Driven Edges}
  Within this syntax, all iterative processes are represented with driver nodes. 
  The driver node shares some basic qualities with the model node. It can have incoming 
  and outgoing function edges to and from variable nodes. For example, many optimizers take 
  convergence tolerance as input, and they output iteration counts and the final objective value. 
  With respect to variable nodes and function edges, drivers behave similarly to
  model nodes and are subject to the same rules governing their use in the graph. In this
  context, driver nodes can be part of driver blocks, which behave identically to analysis 
  blocks.

  Driven edges are the distinguishing characteristic associated with driver nodes and
  driver blocks. Driven edges do not follow the same rules as other edges: they  
   always connect a driver node to a variable node (in either direction), but they 
  are not considered when counting the indegree or outdegree of the variable node.
  Hence, a single variable node could have multiple incoming driven edges from different 
  driver blocks. This accommodates a sequential optimization strategy in which both drivers, at 
  different times, select values for the same variable. Figure \ref{f:driver block} shows 
  a notional example in which a driver block represents an optimizer solving an 
  unconstrained minimization problem.

  \begin{figure}[htb]
    \begin{center}
    \includegraphics[width=3.0in]{images/driver_block}
    \end{center}
    \vspace{-10pt}
  \caption{Example driver block implementation. The driver, analysis, and expression blocks are dash-dotted, dotted, and dashed, respectively. The driver nodes are represented by rotated squares.}
  \label{f:driver block}
  \end{figure}

@@

\subsection{Design Variables, Holes, and Collisions}
  Design variables are input variables whose values are free to be 
  changed by the designer or an optimizer in an MDAO study. In the graph-based 
  syntax, a node corresponds to a design variable if the indegree limits of the node are 
  prescribed as zero.

  If the indegree limits are violated by the number of incoming connection edges 
  (the indegree of the node), then the graph is regarded as an invalid problem formulation. 
  There are two classifications for these violations:
    \begin{description}
      \item[\bf{hole:}] The number of incoming edges is less than the lower indegree limit:
        \begin{equation} 
          \txt{deg}^-(v) < \txt{deg}_l^-(v). 
          \label{e:hole} 
        \end{equation}
        A hole represents a lack of information being supplied to the variable node. 
        This implies that the analysis tool being represented by the analysis block 
        would not be capable of producing outputs.
      \item[\bf{collision:}] The number of incoming edges is greater than the upper 
      indegree limit:
        \begin{equation} 
        \txt{deg}^-(v) > \txt{deg}_u^-(v). 
        \label{e:collision}
        \end{equation}A collision represents redundant information being supplied to 
        the variable node from two or more sources. This conflict implies an 
        ambiguity as to which information is to be used as an input for the analysis tool.
	In multifidelity problems (discussed subsequently), the upper indegree limit could be 
  specified as $\txt{deg}_u^- > 2$, which implies that a collision is not noted in the case of 
  multiple inputs.
  \end{description}

@@

\subsection{Coupling Between Analyses}
  In MDAO, coupling is the mutual dependence between two or more analysis 
  tools and their respective outputs. In our graph-based syntax, coupling is 
  described by the presence of a cycle between two or more analysis blocks. In the Sellar 
  problem from Eq.~\ref{eqn:simple_fpf}, 
  a mutual dependence between $D_1$ and $D_2$ exists through the variables $y_1$ and $y_2$. 
  The edges belonging to this cycle in the Sellar problem are highlighted in 
  Fig. \ref{f:sellar cycles}.  These edges collectively form a closed path between 
  analysis blocks $D_1$ and $D_2$.

@@

  Coupling cycles do not contain driver nodes or driven edges since these 
  are not allowed in the MCG or FPG representations. 
  Solvers, optimizers, and other iterative processes are not involved in the coupling 
  definition; however, one or more solvers will be required to build a valid PSG from a 
  given FPG  that includes cycles.

  Additionally, a coupling cycle has no inherent 
  start or end point. It would be acceptable to select a
  node in the cycle as a starting point and proceed around the
  loop until the starting point is reached again. For the Sellar problem, selecting 
  $D_1$ as the starting point would yield the form of the stated in 
  Eq.~\ref{eqn:simple}.

  Larger problems can involve more complex cycles in their corresponding FPG, indicating more 
  complex coupling between analyses. For example, a cycle could incorporate more than 
   two analysis blocks. Multiple independent cycles could also exist, indicating 
  independent coupling relationships. Cycles could also overlap, meaning that the same analysis 
  blocks belong to multiple different coupling cycles. All of these situations
  arise naturally as problems grow in size, and managing these increasingly complex  couplings is particularly challenging. 
  In the present work, Sec.~\ref{s:example problem}
  demonstrates how building an FPG from an MCG provides an opportunity to 
  identify and potentially reduce the number of cycles in a graph.

@@

\subsection{Multi-fidelity Problems}
  \label{ss:multi-fideliy problems}
  A multi-fidelity MDAO problem can be characterized as a formulation in which 
  two or more different analyses each calculate the same data. Multi-fidelity 
  is represented in a graph by a variable node having an indegree greater than 
  one, which means that multiple connection edges are directed into it. When the 
  upper indegree limit of a variable node is set above one, then the node is 
  implied to allow multiple fidelity instances of the variable. When the lower 
  indegree limit of a variable node is set above one, then multiple fidelity instances are required.

@@

  Multi-fidelity problems are always characterized by graphs in which at least some of the variable 
  nodes have an $\txt{deg}^-_u(v) > 1$. These problems require 
  special techniques for resolving the conflicting edges by introducing some mechanism
  to manage when each of the different fidelity analyses are 
  run \cite{march2012provably,alexandrov2001approximation,Huang_Allen_Notz_Miller_2006}.
  The specifics of this mechanism are not represented within an MCG or 
  FPG. Instead, the multi-fidelity mechanism specifics would be represented as a 
  driver in a PSG derived from a multi-fidelity FPG.

Table \ref{t:variable node classification} summarizes the classification of a variable 
node as a hole, design variable, single valid input (nominal), collision, or 
multi-fidelity node in relation to the indegree limits.

\begin{table}[htbp]
  \centering
  \caption{Variable Node Classification}
    \begin{tabular}{ccccc}
    \toprule
    $\txt{deg}^-_l(v)$ & $\txt{deg}^-_u(v)$ & $\txt{deg}^-(v)$ & valid & classification \\
    \midrule
    0     & 0     & 0     & yes   & design variable \\
    0     & 0     & 1     & no   & collision \\
    0     & 1     & 0     & yes   & design variable \\
    0     & 1     & 1     & yes   & single input \\
    1     & 1     & 0     & no    & hole \\
    1     & 1     & 1     & yes   & single input \\
    1     & 1     & $>1$ & no    & collision \\
    1     & 2     & 1     & yes   & supplied input \\
    1     & 2     & 2     & yes   & multi-fidelity \\
    2     & 2     & $<2$ & no    & hole \\
    2     & 2     & 2     & yes   & multi-fidelity \\
    2     & 2     & $>2$     & no   & collision \\
    \bottomrule
    \end{tabular}%
  \label{t:variable node classification}%
\end{table}%

\subsection{Iterative Control: Drivers}
  There are many iterative processes  such as 
  solvers, optimizers, and designs of experiments that are used to solve MDAO problems. 
  These processes represent the control structures that govern the path toward a solution, but they do not fundamentally impact the definition of the 
  problem to be solved. Hence, driver 
  nodes and driven edges are excluded from both the MCG and FPG, which deal exclusively 
  with problem formulation. These nodes and edges are used only in the PSG, in which the issue of 
  how to solve a specific problem is addressed. Indeed, at least one driver block is 
  required to be present in any PSG in order to select values for the design variables during the optimization process.
